Overt attentional capture by reward-related stimuli overcomes inhibitory suppression

### 2019-04-18 analysis/paper/paper.tex
update manuscript

\documentclass[jou, a4paper, noextraspace,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Overt attentional capture by reward-related stimuli overcomes top-down suppression},
            pdfauthor={Daniel Pearson, Poppy Watson, Phillip Cheng, \& Mike E. Le Pelley},
            pdfkeywords={attentional capture, reward, suppression, cognitive control},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Overt attentional capture by reward-related stimuli overcomes top-down
suppression}
    \author{Daniel Pearson\textsuperscript{1}, Poppy Watson\textsuperscript{1},
Phillip Cheng\textsuperscript{2}, \& Mike E. Le
Pelley\textsuperscript{1}}
    \date{}

@@

Recently, a hybrid model of attentional selection has been
proposed---the \emph{signal suppression hypothesis} (Gaspelin \& Luck,
2018b; Sawaki \& Luck, 2010)---which attempts to integrate these two
competing accounts. According to this model, salient stimuli
automatically generate an attentional priority signal (or
\enquote{attend-to-me} signal) that will go on to capture attention
unless it is suppressed by a top-down control process, which builds over
repeated experience with the salient stimulus (Gaspelin \& Luck, 2018a).
This account is consistent with stimulus-driven theories, in that it
proposes that physically salient stimuli have the potential to
involuntarily capture attention; however it is also consistent with
goal-directed theories in that it proposes that there is some top-down
control of the allocation of attention. Evidence supporting the signal
suppression hypothesis has largely come from the event-related potential
(ERP) literature (for review, see Gaspelin \& Luck, 2019), however
recent studies have demonstrated behavioural evidence of suppression in
the context of overt attention (Gaspelin \& Luck, 2018a; i.e.,
eye-movements: see Gaspelin, Leonard, \& Luck, 2017; Ipata, Gee,
Gottlieb, Bisley, \& Goldberg, 2006) and covert attention (i.e., shifts
of attention that are not accompanied by eye-movements: see Gaspelin,
Leonard, \& Luck, 2015).

For example, across two experiments Gaspelin et al. (2017) had
participants perform variants of a commonly used procedure for examining
attentional capture---the \emph{additional singleton task} (J. Theeuwes,
1991, 1992)---while recording eye movements. Eye movements provide a
useful index of attention since an eye movement to a location (a
saccade) is necessarily preceded by a shift of spatial attention to that
location (Deubel \& Schneider, 1996). In this task, participants were
required to locate and respond to a shape-singleton target that was
presented in an array of non-target shapes. On half of the trials, one
of the non-target shapes was rendered in a different colour from the
rest of the array so as to be a physically salient colour-singleton
distractor. If participants were \emph{more} likely to make a saccade
towards the singleton distractor than one of the nonsingleton
non-targets, this would indicate that the singleton distractor had
captured attention. By contrast, if participants were \emph{less} likely
to make a saccade to the singleton distractor than one of the
nonsingleton non-targets, this would suggest that attention to the
singleton distractor had been suppressed. In Gaspelin et al. (2017)'s
Experiment 1, participants were encouraged to adopt a strategy of
searching for the \enquote{odd-one-out} in the display (i.e,
\emph{singleton-search mode}; Bacon \& Egeth, 1994), by having the
target be randomly either a circle amongst diamonds or a diamond amongst
circles on each trial (see Figure \ref{fig:methodFig}A for an example
from the current study). Under these conditions, both stimulus-driven
(J. Theeuwes, 1992) and goal-directed (Bacon \& Egeth, 1994) accounts of
attentional selection would predict capture, as the template that
participants use to locate the target (i.e., any singleton) is broad
enough to include the singleton distractor. As expected, under these
singleton-search conditions Gaspelin et al. found that participants'
overt attention was captured by the salient singleton. In Experiment 2,
participants were instead encouraged to adopt a more specific target
template (i.e., \emph{feature-search mode}; Bacon \& Egeth, 1994), by
consistently mapping the target identity to one shape (e.g., search for
a circle on every trial), and having a heterogeneous set of shapes act
as the non-targets (see Figure \ref{fig:methodFig}B). This manipulation
makes searching for any singleton a highly inefficient strategy, as
there are multiple singleton nontargets in the display (e.g., a circle,
a square, a triangle, and a red shape). Thus, participants should be
encouraged to search for the specific features that define the target
(i.e., the \enquote{circleness} of the circle). Under these conditions,
attention to the salient colour-singleton distractor was suppressed
relative to the other non-targets, suggesting that it was treated
differently from the other non-targets (inconsistent with goal-directed
accounts), yet did not capture gaze (inconsistent with stimulus-driven
accounts). By contrast, the signal suppression hypothesis can account
for these findings by suggesting that the singleton generated an
attentional priority signal as a consequence of its physical salience,
but this signal could be suppressed by top-down control under conditions
that promote a more specific search strategy (i.e., feature search).
Importantly, this suppression effect was observed even for the most
rapidly generated eye movements, which was taken as evidence that the
suppression effect operates early in processing, before attention is
initially allocated.

@@

Pairing a stimulus with reward is thought to increase its attentional
priority in a way that mimics or otherwise interacts with its physical
salience. For example, eye-tracking studies have demonstrated that VMAC
is largest for the most rapidly generated saccades, and diminishes with
increasing saccade latency (Failing, Nissens, Pearson, Le Pelley, \&
Theeuwes, 2015; Pearson et al., 2016), a pattern similar to that
observed in capture by physical salience (e.g., Zoest, Donk, \&
Theeuwes, 2004). Similarly, electrophysiological studies have shown that
reward affects ERP components thought to index low-level perceptual
activity (Hickey, Chelazzi, \& Theeuwes, 2010). This is consistent with
the idea of \emph{incentive salience} (Berridge \& Robinson, 1998),
wherein a reward-related stimulus becomes more salient and attention
grabbing to the observer through a change in its perceptual
representation. This begs the question, if associating a stimulus with
reward increases its perceptual priority as though it has become more
physically salient, can this priority signal be suppressed by the
top-down control process proposed by the signal suppression hypothesis?
That is, can incentive salience be suppressed in the same way as
physical salience? Aside from the theoretical importance of this
question, it may also have important clinical implications. Recent
evidence suggests that the processes underlying VMAC are closely related
to the attentional biases towards drug-related stimuli observed in
addiction (Albertella et al., 2017; Brian A Anderson, Faulkner, Rilee,
Yantis, \& Marvel, 2013; for review see Field \& Cox, 2008), which can
promote relapse in recovering addicts (Marhe, Waters, Wetering, \&
Franken, 2013; Marissen et al., 2006; A. J. Waters, Marhe, \& Franken,
2012). Thus it is important to know whether top-down cognitive control
processes can be used to overcome the influence of reward on attention
when such capture is counterproductive and contrary to our goals. We
addressed this question by combining Le Pelley et al.'s VMAC procedure
with Gaspelin et al.'s (2017) overt attentional suppression procedure to
investigate whether reward-related attentional biases can be suppressed
by top-down control processes under conditions promoting feature search.

@@

The task was based on that used by Gaspelin et al. (2017) and Pearson et
al. (2015, 2016). Each trial consisted of a fixation display, a search
display, and a feedback display (see Figure \ref{fig:methodFig}C). All
stimuli were presented on a black background. The fixation display
consisted of a white cross (0.5\(^\circ\) visual angle), presented in
the centre of the screen and surrounded by a white circle (diameter
3.0\(^\circ\)). The search display consisted of four filled shapes (size
1.6\(^\circ\) \(\times\) 1.6\(^\circ\)) presented above, below, left and
right of screen centre (eccentricity 4.5\(^\circ\)). One of the shapes
was designated as the target, and the other three shapes were designated
non-targets. For all participants, the target was randomly determined to
be either a circle or a diamond on each trial.\footnote{This aspect of
  our design differs from that of Gaspelin et al. (2017). In their
  study, for half of participants in the feature search condition the
  target was always a circle, and for other participants the target was
  always a diamond. By contrast, for the singleton search condition, for
  all participants the target was a circle among diamonds on some
  trials, and a diamond among circles on others. That is, target
  identity (circle/diamond) was manipulated between-subjects for the
  feature search condition, and within-subjects for the singleton search
  condition. In the current study, we equated the two conditions by
  varying target identity within-subjects for all participants. This
  ensured that any difference between the two search conditions was
  driven by a difference in the specificity of participants' target
  template and not by a difference in participants' history of selecting
  the target shape (e.g., Maljkovic \& Nakayama, 1994).} The identity of
the non-target shapes varied by search condition. For the
\emph{singleton search} condition, the non-target stimuli were all
rendered as the alternative target shape (i.e., if the target was a
diamond, all of the non-targets were circles; Figure
\ref{fig:methodFig}B). For the \emph{feature search} condition, the
non-target stimuli were randomly selected without replacement from a set
of two squares and two hexagons, and were never rendered as the
alternative target shape (Figure \ref{fig:methodFig}C). The target and
two of the non-targets were always rendered in grey (CIE \(x\),\(y\)
chromaticity coordinates of .327/.400). On the majority of trials, one
of the non-target shapes (the \emph{singleton distractor}) was rendered
in either orange or blue (CIE \(x\),\(y\) coordinates .492/.445 and
.192/.216, respectively). On the remaining trials, all shapes were
rendered in grey such that there was no singleton distractor. The values
of blue and orange had similar luminance (\textasciitilde{}24.5
cd/m\textsuperscript{2}), which was higher than that of grey
(\textasciitilde{}8.3 cd/m\textsuperscript{2}). The feedback display
showed the points earned on the current trial. If the response time (RT)
exceeded the soft-timeout threshold (see below), the message
\enquote{Too slow} appeared below the reward feedback.

@@

Figure \ref{fig:SaccadePlot} shows the percentage of first saccades
directed towards each stimulus type (i.e., the target\footnote{Note that
  the comparison of interest in determining whether a stimulus captured
  attention, or was otherwise suppressed, is the percentage of first
  saccades to the singleton vs.~the average nonsingleton distractor.
  Thus, while we have included the percentage of first saccades to the
  target in Figure \ref{fig:SaccadePlot} for the sake of completeness,
  we have not included these data in our analyses.}, average
nonsingleton distractor, and singleton distractor) on high- and
low-reward trials for the singleton search (Figure
\ref{fig:SaccadePlot}A) and feature search (Figure
\ref{fig:SaccadePlot}C) conditions; Figures \ref{fig:SaccadePlot}B and
\ref{fig:SaccadePlot}D show capture scores (difference in percentage of
first saccades directed towards singleton versus average nonsingleton
distractor) for the singleton search and feature search conditions,
respectively. ANOVA analysis of capture scores revealed a significant
main effect of trial type (high-reward vs low-reward),
\(F(1, 62) = 45.55\), \(p < .001\), \(\hat{\eta}^2_p = .424\), with the
singleton capturing attention more often on high-reward trials than
low-reward trials; once again demonstrating a VMAC effect. A main effect
of search condition (singleton search vs feature search),
\(F(1, 62) = 4.89\), \(p = .031\), \(\hat{\eta}^2_p = .073\), indicated
that both the high-reward and low-reward singleton captured attention
more often, on average, in the singleton search condition than the
feature search condition. Importantly, the trial type \(\times\) search
condition interaction was non-significant, \(F(1, 62) = 0.12\),
\(p = .735\), \(\hat{\eta}^2_p = .002\), with a corresponding Bayes
factor of \(\mathrm{BF}_{\textrm{01}} = 3.73\) indicating moderate
support for the null hypothesis that search condition had no effect on
the magnitude of the VMAC effect on first saccade direction.

Next we tested whether overt attention to each distractor type was
captured or suppressed in each search condition. Planned one-sample
\emph{t}-tests showed that capture scores were significantly greater
than zero (indicating that overt attention was captured by the
high-reward singleton) under conditions promoting singleton search,
\(t(31) = 9.03\), \(p < .001\), \(d_z=\) 1.60, as well as feature
search, \(t(31) = 3.98\), \(p < .001\), \(d_z=\) 0.70. For the
low-reward distractor, there was a non-significant numerical trend
towards capture in the singleton search condition, \(t(31) = 1.02\),
\(p = .313\), \(d_z=\) 0.18. In contrast, for the feature search
condition, capture score was significantly below zero, \(t(31) = 2.30\),
\(p = .028\), \(d_z=\) 0.41, indicating that overt attention to the
low-reward singleton was suppressed. Thus, in line with previous
findings (Gaspelin et al., 2017), our data suggest that overt attention
to physically salient singletons can be suppressed when participants are
encouraged to adopt a specific target template. However, singletons that
have been associated with high-reward reward capture overt attention,
regardless of participants' target template and despite such capture
being directly counterproductive to earning reward.

@@

To investigate the time course of the overt attentional capture and
suppression effects, we analysed capture scores (proportion of first
saccades towards the singleton distractor minus the average nonsingleton
distractor) on high-reward and low-reward trials as a function of
saccade latency using the Vincentizing procedure (Ratcliff, 1979). We
calculated mean saccade latency and capture score separately for each
trial type and quartile of the individual saccade latency distributions.
We then analysed the capture scores for the most rapidly generated
saccades. According to the signal suppression hypothesis, the top-down
suppression effect is activated early in processing, prior to the
initial allocation of attention (Gaspelin \& Luck, 2019). Therefore the
speed with which a saccade is initiated should have no effect on the
extent to which overt attention can be suppressed. As illustrated in
Figure \ref{fig:VincentizedPlot}, the most rapidly generated saccades
were captured by the high-reward singleton (capture scores \(>0\)) in
both the singleton search, \(t(31) = 13.02\), \(p < .001\), \(d_z=\)
2.30, and feature search conditions, \(t(31) = 6.45\), \(p < .001\),
\(d_z=\) 1.14. Similarly, the the most rapidly generated saccades were
captured by the low-reward singleton in both the singleton search,
\(t(31) = 4.10\), \(p < .001\), \(d_z=\) 0.72, and feature search
conditions, \(t(31) = 3.34\), \(p = .002\), \(d_z=\) 0.59. It is notable
that we observed capture (rather than suppression) for the fastest
saccades for both trial-types in the feature search condition. This
finding is inconsistent with Gaspelin et al.'s (2017) previous report of
attentional suppression of physically salient singleton distractors in
the fastest eye-movements.

>>>>

\documentclass[jou, a4paper, noextraspace,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Overt attentional capture by reward-related stimuli overcomes inhibitory suppression},
            pdfauthor={Daniel Pearson, Poppy Watson, Phillip Cheng, \& Mike E. Le Pelley},
            pdfkeywords={attentional capture, reward, suppression, cognitive control},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Overt attentional capture by reward-related stimuli overcomes inhibitory
suppression}
    \author{Daniel Pearson\textsuperscript{1}, Poppy Watson\textsuperscript{1},
Phillip Cheng\textsuperscript{2}, \& Mike E. Le
Pelley\textsuperscript{1}}
    \date{}

@@

Recently, a hybrid model of attentional selection has been
proposed---the \emph{signal suppression hypothesis} (Gaspelin \& Luck,
2018b; Sawaki \& Luck, 2010)---which attempts to integrate these two
competing accounts. According to this model, salient stimuli
automatically generate an attentional priority signal (or
\enquote{attend-to-me} signal) that will go on to capture attention
unless it is suppressed by an inhibitory control process, which builds
over repeated experience with the salient stimulus (Gaspelin \& Luck,
2018a). This account is consistent with stimulus-driven theories, in
that it proposes that physically salient stimuli have the potential to
involuntarily capture attention; however it is also consistent with
goal-directed theories in that it proposes that there is some top-down
control of the allocation of attention. Evidence supporting the signal
suppression hypothesis has largely come from the event-related potential
(ERP) literature (for review, see Gaspelin \& Luck, 2019), however
recent studies have demonstrated behavioural evidence of suppression in
the context of overt attention (Gaspelin \& Luck, 2018a; i.e.,
eye-movements: see Gaspelin, Leonard, \& Luck, 2017; Ipata, Gee,
Gottlieb, Bisley, \& Goldberg, 2006) and covert attention (i.e., shifts
of attention that are not accompanied by eye-movements: see Gaspelin,
Leonard, \& Luck, 2015).

For example, across two experiments Gaspelin et al. (2017) had
participants perform variants of a commonly used procedure for examining
attentional capture---the \emph{additional singleton task} (J. Theeuwes,
1991, 1992)---while recording eye movements. Eye movements provide a
useful index of attention since an eye movement to a location (a
saccade) is necessarily preceded by a shift of spatial attention to that
location (Deubel \& Schneider, 1996). In this task, participants were
required to locate and respond to a shape-singleton target that was
presented in an array of non-target shapes. On half of the trials, one
of the non-target shapes was rendered in a different colour from the
rest of the array so as to be a physically salient colour-singleton
distractor. If participants were \emph{more} likely to make a saccade
towards the singleton distractor than one of the nonsingleton
non-targets, this would indicate that the singleton distractor had
captured attention. By contrast, if participants were \emph{less} likely
to make a saccade to the singleton distractor than one of the
nonsingleton non-targets, this would suggest that attention to the
singleton distractor had been suppressed. In Gaspelin et al. (2017)'s
Experiment 1, participants were encouraged to adopt a strategy of
searching for the \enquote{odd-one-out} in the display (i.e,
\emph{singleton-search mode}; Bacon \& Egeth, 1994), by having the
target be randomly either a circle amongst diamonds or a diamond amongst
circles on each trial (see Figure \ref{fig:methodFig}A for an example
from the current study). Under these conditions, both stimulus-driven
(J. Theeuwes, 1992) and goal-directed (Bacon \& Egeth, 1994) accounts of
attentional selection would predict capture, as the template that
participants use to locate the target (i.e., any singleton) is broad
enough to include the singleton distractor. As expected, under these
singleton-search conditions Gaspelin et al. found that participants'
overt attention was captured by the salient singleton. In Experiment 2,
participants were instead encouraged to adopt a more specific target
template (i.e., \emph{feature-search mode}; Bacon \& Egeth, 1994), by
consistently mapping the target identity to one shape (e.g., search for
a circle on every trial), and having a heterogeneous set of shapes act
as the non-targets (see Figure \ref{fig:methodFig}B). This manipulation
makes searching for any singleton a highly inefficient strategy, as
there are multiple singleton nontargets in the display (e.g., a circle,
a square, a triangle, and a red shape). Thus, participants should be
encouraged to search for the specific features that define the target
(i.e., the \enquote{circleness} of the circle). Under these conditions,
attention to the salient colour-singleton distractor was suppressed
relative to the other non-targets, suggesting that the visual attention
system treated the salient singleton differently from the other
non-targets (inconsistent with goal-directed accounts), yet did not
capture gaze (inconsistent with stimulus-driven accounts). By contrast,
the signal suppression hypothesis can account for these findings by
suggesting that the singleton generated an attentional priority signal
as a consequence of its physical salience, but this signal could be
suppressed by an inhibitory mechanism under conditions that promoted a
more specific search strategy (i.e., feature search). Importantly, this
suppression effect was observed even for the most rapidly generated eye
movements, which was taken as evidence that the suppression effect
operates early in processing, before attention is initially allocated.

@@

Pairing a stimulus with reward is thought to increase its attentional
priority in a way that mimics or otherwise interacts with its physical
salience. For example, eye-tracking studies have demonstrated that VMAC
is largest for the most rapidly generated saccades, and diminishes with
increasing saccade latency (Failing, Nissens, Pearson, Le Pelley, \&
Theeuwes, 2015; Pearson et al., 2016), a pattern similar to that
observed in capture by physical salience (e.g., Zoest, Donk, \&
Theeuwes, 2004). Similarly, electrophysiological studies have shown that
reward affects ERP components thought to index low-level perceptual
activity (Hickey, Chelazzi, \& Theeuwes, 2010). This is consistent with
the idea of \emph{incentive salience} (Berridge \& Robinson, 1998),
wherein a reward-related stimulus becomes more salient and attention
grabbing to the observer through a change in its perceptual
representation. This raises the question, if associating a stimulus with
reward increases its perceptual priority as though it has become more
physically salient, can this priority signal be suppressed by the
inhibitory control process proposed by the signal suppression
hypothesis? That is, can incentive salience be suppressed in the same
way as physical salience? Aside from the theoretical importance of this
question, it may also have important clinical implications. Recent
evidence suggests that the processes underlying VMAC are closely related
to the attentional biases towards drug-related stimuli observed in
addiction (Albertella et al., 2017; Brian A Anderson, Faulkner, Rilee,
Yantis, \& Marvel, 2013; for review see Field \& Cox, 2008), which can
promote relapse in recovering addicts (Marhe, Waters, Wetering, \&
Franken, 2013; Marissen et al., 2006; A. J. Waters, Marhe, \& Franken,
2012). Thus it is important to know whether cognitive control processes
can be used to overcome the influence of reward on attention when such
capture is counterproductive and contrary to our goals. We addressed
this question by combining Le Pelley et al.'s VMAC procedure with
Gaspelin et al.'s (2017) overt attentional suppression procedure to
investigate whether reward-related attentional biases can be suppressed
by inhibitory control processes under conditions promoting feature
search.

@@

The task was based on that used by Gaspelin et al. (2017) and Pearson et
al. (2015, 2016). Each trial consisted of a fixation display, a search
display, and a feedback display (see Figure \ref{fig:methodFig}C). All
stimuli were presented on a black background. The fixation display
consisted of a white cross (0.5\(^\circ\) visual angle), presented in
the centre of the screen and surrounded by a white circle (diameter
3.0\(^\circ\)). The search display consisted of four filled shapes (size
1.6\(^\circ\) \(\times\) 1.6\(^\circ\)) presented above, below, left and
right of screen centre (eccentricity 4.5\(^\circ\)). One of the shapes
was designated as the target, and the other three shapes were designated
non-targets. For all participants, the target was randomly determined to
be either a circle or a diamond on each trial.\footnote{This aspect of
  our design differs from that of Gaspelin et al. (2017). In their
  study, for half of participants in the feature search condition the
  target was always a circle, and for other participants the target was
  always a diamond. By contrast, for the singleton search condition, for
  all participants the target was a circle among diamonds on some
  trials, and a diamond among circles on others. That is, target
  identity (circle/diamond) was manipulated between-subjects for the
  feature search condition, and within-subjects for the singleton search
  condition. In the current study, we equated the two conditions by
  varying target identity within-subjects for all participants. This
  ensured that any difference between the two search conditions was
  driven by a difference in the specificity of participants' target
  template and not by a difference in participants' history of selecting
  the target shape (e.g., Maljkovic \& Nakayama, 1994).} The identity of
the non-target shapes varied by search condition. For the
\emph{singleton search} condition, the non-target stimuli were all
rendered as the alternative target shape (i.e., if the target was a
diamond, all of the non-targets were circles; Figure
\ref{fig:methodFig}B). For the \emph{feature search} condition, the
non-target stimuli were randomly selected without replacement from a set
of two squares and two hexagons (such that one of the non-targets was
always a shape singleton), and were never rendered as the alternative
target shape (Figure \ref{fig:methodFig}C). The target and two of the
non-targets were always rendered in grey (CIE \(x\),\(y\) chromaticity
coordinates of .327/.400). On the majority of trials, one of the
non-target shapes (the \emph{singleton distractor}) was rendered in
either orange or blue (CIE \(x\),\(y\) coordinates .492/.445 and
.192/.216, respectively). On the remaining trials, all shapes were
rendered in grey such that there was no singleton distractor. The values
of blue and orange had similar luminance (\textasciitilde{}24.5
cd/m\textsuperscript{2}), which was higher than that of grey
(\textasciitilde{}8.3 cd/m\textsuperscript{2}). The feedback display
showed the points earned on the current trial. If the response time (RT)
exceeded the soft-timeout threshold (see below), the message
\enquote{Too slow} appeared below the reward feedback.

@@

Figure \ref{fig:SaccadePlot} shows the percentage of first saccades
directed towards each stimulus type (i.e., the target\footnote{Note that
  the comparison of interest in determining whether a stimulus captured
  attention, or was suppressed, is the percentage of first saccades to
  the singleton vs.~the average nonsingleton distractor. Thus, while we
  have included the percentage of first saccades to the target in Figure
  \ref{fig:SaccadePlot} for the sake of completeness, we have not
  included these data in our analyses.}, average nonsingleton
distractor, and singleton distractor) on high- and low-reward trials for
the singleton search (Figure \ref{fig:SaccadePlot}A) and feature search
(Figure \ref{fig:SaccadePlot}C) conditions; Figures
\ref{fig:SaccadePlot}B and \ref{fig:SaccadePlot}D show capture scores
(difference in percentage of first saccades directed towards singleton
versus average nonsingleton distractor) for the singleton search and
feature search conditions, respectively. ANOVA analysis of capture
scores revealed a significant main effect of trial type (high-reward vs
low-reward), \(F(1, 62) = 45.55\), \(p < .001\),
\(\hat{\eta}^2_p = .424\), with the singleton capturing attention more
often on high-reward trials than low-reward trials; once again
demonstrating a VMAC effect. A main effect of search condition
(singleton search vs feature search), \(F(1, 62) = 4.89\), \(p = .031\),
\(\hat{\eta}^2_p = .073\), indicated that both the high-reward and
low-reward singleton captured attention more often, on average, in the
singleton search condition than the feature search condition.
Importantly, the trial type \(\times\) search condition interaction was
non-significant, \(F(1, 62) = 0.12\), \(p = .735\),
\(\hat{\eta}^2_p = .002\), with a corresponding Bayes factor of
\(\mathrm{BF}_{\textrm{01}} = 3.73\) indicating moderate support for the
null hypothesis that search condition had no effect on the magnitude of
the VMAC effect on first saccade direction.

Next we tested whether overt attention to each distractor type was
captured or suppressed in each search condition. Planned one-sample
\emph{t}-tests showed that capture scores were significantly greater
than zero (indicating that overt attention was captured by the
high-reward singleton) under conditions promoting singleton search,
\(t(31) = 9.03\), \(p < .001\), \(d_z=\) 1.60, as well as feature
search, \(t(31) = 3.98\), \(p < .001\), \(d_z=\) 0.70. For the
low-reward distractor, there was a non-significant numerical trend
towards capture in the singleton search condition, \(t(31) = 1.02\),
\(p = .313\), \(d_z=\) 0.18. In contrast, for the feature search
condition, capture score was significantly below zero, \(t(31) = 2.30\),
\(p = .028\), \(d_z=\) 0.41, indicating that overt attention to the
low-reward singleton was suppressed. Thus, in line with previous
findings (Gaspelin et al., 2017), our data suggest that overt attention
to physically salient singletons can be suppressed when participants are
encouraged to adopt a specific target template. However, singletons that
have been associated with high reward capture overt attention,
regardless of participants' target template and despite such capture
being directly counterproductive to earning reward.

@@

To investigate the time course of the overt attentional capture and
suppression effects, we analysed capture scores (proportion of first
saccades towards the singleton distractor minus the average nonsingleton
distractor) on high-reward and low-reward trials as a function of
saccade latency using the Vincentizing procedure (Ratcliff, 1979). We
calculated mean saccade latency and capture score separately for each
trial type and quartile of the individual saccade latency distributions.
We then analysed the capture scores for the data from the first quartile
of this distribution, i.e., the most rapidly generated saccades.
According to the signal suppression hypothesis, the attentional
suppression effect is activated early in processing, prior to the
initial allocation of attention (Gaspelin \& Luck, 2019). Therefore the
speed with which a saccade is initiated should have no effect on the
extent to which overt attention can be suppressed. As illustrated in
Figure \ref{fig:VincentizedPlot}, the most rapidly generated saccades
were captured by the high-reward singleton (capture scores \(>0\)) in
both the singleton search, \(t(31) = 13.02\), \(p < .001\), \(d_z=\)
2.30, and feature search conditions, \(t(31) = 6.45\), \(p < .001\),
\(d_z=\) 1.14. Similarly, the most rapidly generated saccades were
captured by the low-reward singleton in both the singleton search,
\(t(31) = 4.10\), \(p < .001\), \(d_z=\) 0.72, and feature search
conditions, \(t(31) = 3.34\), \(p = .002\), \(d_z=\) 0.59. It is notable
that we observed capture (rather than suppression) for the fastest
saccades for both trial-types in the feature search condition. This
finding is inconsistent with Gaspelin et al.'s (2017) previous report of
attentional suppression of physically salient singleton distractors in
the fastest eye-movements.
